Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-06-26 manuscript.qmd
adjust mid ext theory section

For later use, we define at this point the matrices $\boldsymbol{\Lambda}$:

@@

The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-small-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors. In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories. Therefore, our parameter space is:

@@

Together with the explicit modelling of the tumor's midline extension $\epsilon$ from @sec-midline, we now have a model that may be capable of capturing the higher prevalence of contralateral involvment that comes with tumors extending over the mid-sagittal line.

@@

Technically, the introduction of this additional random variable doubles the space of the hidden states and therefore quadruples the size of the transition matrix $\mathbf{A}$. But since only the transition matrix in case of midline extension depends on the lateralization of the tumor, we implement it differently and thereby make the computation slightly more efficient.

First, the overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are respectively given by

@@

Using this, it is straightforward to write down the joint probability of ipsi- and contralalteral states in case of a tumor that does *not* extend over the mid-sagittal line similar to the marginalization in @eq-bilateral-marginal:

$$
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{False} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot (1 - p_\epsilon)^t \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t, \epsilon=\texttt{False} \right)
$$ {#eq-joint-noext}

where for the contralateral evolution we used the transition matrix that depends on the base spread parameters $b_v^{\text{c},\epsilon=\texttt{False}}$.

The case when midline extension is eventually present is more complicated: We already marginalize over the exact time-step when the tumor grows over the mid-sagittal line. But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates from the tumor to the contralateral LNLs, given by the linear mix in @eq-mixing. We can achieve the correct marginalization by iteratively building the joint $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$ from $t=0$, where we know that all LNLs are healthy and the tumor is lateralized:

$$
P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
$$

@@

1. The tumor was still lateralized at $t=\tau$ and just grew over the midline. Under this scenario, the probability for having a midline extension at $t=\tau+1$ is $p_\epsilon$ and we use this to weight the contralateral state distribution that was so far evolved without increased contralateral spread.
2. The mid-sagittal line was already crossed by the tumor. In that case, the probability is 1 to remain in that lateralization state. Thus, for this case we simply add the same distribution we are computing, but from one time-step earlier, arriving at a recursion relation:

$$
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}^{\text{c},\epsilon=\texttt{True}}
$$

Using this, we can write the joint of ipsi- and contralateral involvement also for the case of mid-sagittal extension:

$$
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)
$$ {#eq-joint-ext}

Subsequently, @eq-joint-noext and @eq-joint-ext can be used to compute the likelihoods of patients with the respective tumor lateralization. If, for some reason, it is not known whether the tumor crossed the midline or not, we can simply add the two distributions to marginalize over midline extension as well.

In @fig-model-state-dist we plot the state distribution for a full midline model as two separate heatmaps. To keep it readable, this example model only considers the LNLs II, III, and IV ipsi- and contralaterally.

![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.](figures/fig-model-state-dist.png){#fig-model-state-dist}

>>>>

For later use, we define at this point the matrices of the dsitributions over all hidden states, given all time-steps $\boldsymbol{\Lambda}$:

@@

The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-small-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors. In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories. Therefore, the parameter space of the unilateral model is:

@@

Our parameter space has now expanded to

$$
\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.} \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
$$ {#eq-bi-param-space}

which is less than double in size, compared to the unilateral model. From these parameters, we now build three different transition matrices: The unchanged $\mathbf{A}_\text{i}$ for the ipsilateral side, one $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for the contralateral side that covers the progression as long as the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ in case of a tumor that has crossed the mid-sagittal line.

@@

The overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then respectively given by

@@

Using this, it is straightforward to write down the matrix of state distributions for all time-steps, as in @eq-lambda-matrix covering the contralateral hidden state evolution:

$$
\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{False}} =
\left(
\begin{array}{r}
\boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{0\phantom{t_\text{max}}} \\
(1-p_\epsilon) \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{1\phantom{t_\text{max}}} \\
\hfill \vdots \hfill \\
(1-p_\epsilon)^{t_\text{max}} \cdot \boldsymbol{\pi}^\intercal \cdot \left( \mathbf{A}_\text{c}^{\epsilon=\texttt{False}} \right)^{t_\text{max}\phantom{0}} \\
\end{array}
\right)
$$

where we used the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ that depends on the base spread parameters $b_v^{\text{c},\epsilon=\texttt{False}}$.

The case when midline extension is eventually present is more complicated: We already marginalize over the exact time-step when the tumor grows over the mid-sagittal line. But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates $b_v^{\text{c}, \epsilon=\texttt{True}}$ from the tumor to the contralateral LNLs, given by the linear mix in @eq-mixing. We can achieve the correct marginalization by iteratively building the joint $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$. We start from $t=0$, where we know that all LNLs are healthy - i.e., the contralateral neck is in the state $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$ - and the tumor is lateralized ($\epsilon=\texttt{False}$):

$$
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid t=0 \right) = \boldsymbol{\xi}_0
$$

@@

1. The tumor was still lateralized at $t=\tau$ and just grew over the midline. Under this scenario, the probability for having a midline extension at $t=\tau+1$ is given by $p_\epsilon$. We use this to weight the contralateral state distribution that was so far evolved without increased contralateral spread.
2. The mid-sagittal line was already crossed by the tumor. In that case, the probability is 1 to remain in that lateralization state. Thus, for this case we simply add the same distribution we are computing, but from one time-step earlier, arriving at a recursion relation:

$$
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}_\text{c}^{\epsilon=\texttt{True}}
$$

We can collect the iteratively computed distributions for the midline extension case to define the matrix over the states given all time-steps, in analogy to @eq-lambda-matrix:

$$
\boldsymbol{\Lambda}_\text{c}^{\epsilon=\texttt{True}} = \begin{pmatrix}
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 0 \right) \\
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid 1 \right) \\
\vdots \\
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t_\text{max} \right) \\
\end{pmatrix}
$$

Using this, we can again write the joint of ipsi- and contralateral involvement - now also for the case of mid-sagittal extension - algebraically as before in @eq-bilateral-marginal-algebra:

$$
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
$$

We can use this term to compute the likelihood of all patients with and without midline extension separately. And if for some patients the information of tumor lateralization is not available, we can simply sum the above term once for $\epsilon=\texttt{False}$ and once for $\epsilon=\texttt{True}$ to marginalize over the unknown variable.

The final parameter space of our extended model has now reached this size:

$$
\boldsymbol{\theta} = \left( \left\{ b_v^\text{i} \right\}, \left\{ b_v^\text{c} \right\}, \alpha, \left\{ t_{vr} \right\}, p_\text{early}, p_\text{adv.}, p_\epsilon \right) \quad \text{with} \quad \genfrac{}{}{0pt}{2}{v\leq V}{r\in\operatorname{pa}(v)}
$$ {#eq-ext-param-space}

In @fig-model-state-dist we plot the state distribution for a full midline model as two separate heatmaps. To keep it manageable to interpret, this example model only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2 \times 2^3 \times 2^3 = 128$ distinct states.

![3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck. Note the higher probability for contralateral involvement in the right subplot, where the tumor extends over the mid-sagittal line.](figures/fig-model-state-dist.png){#fig-model-state-dist}
